A Test of the Frequency Independence Assumption of Power System Coefficients used in Geomagnetically Induced Current Estimates

### 2019-07-02 paper.tex
Updates from Overleaf

Motivated by the results described in the following section in which Model~2 provides a significantly better representation of GIC than Model~1, we have also considered two additional modeling approaches that can be used to specify GIC. In these two models, a direct estimate of the geoelectric field is not needed; instead, the driver is the geomagnetic field, which is much more frequently measured, and as a result, these models may be of practical interest.

@@

Estimates of $a(\omega)$ and $b(\omega)$ are made at evaluation frequencies, $\omega_e \equiv 2\pi f_e$, by performing a ordinary least squares regression (OLS) using Equation~\ref{model2} and a set $G_E(\omega)$ and $\mathbf{E}(\omega)$ in a band of frequencies around $f_e$ from the DFT (discrete Fourier transform) of 24 hours of 1-s cadence data. Following the suggestion of \cite{Simpson2005}, the 28 values of $f_e$ start at $0.25$ Hz and decrease by a factor of approximately $\sqrt{2}$ and end at $1/43200$~Hz; if a $f_e$ value computed in this way is not an integer multiple of $1/86400$, the nearest smaller integer multiple is used. The periods associated with the evaluation frequencies are listed in Table~\ref{evaluationperiods}. The regression for each $f_e$ uses all DFT frequencies with values in a window with a width of twice the separation from the next lower value of $f_e$ and the given $f_e$ with the exceptions that the lowest evaluation frequency window has a lower bound of $1/86400$~Hz and highest evaluation frequency window has an upper bound of $0.375$~Hz.

@@

Several modifications to this basic procedure were considered. Robust regression \citep{Egbert1986} with a Huber weighting constant of $1.345$ and an upper limit of 50 iterations and bi-square weighting with a constant of $4.685$ (both constants provide 95\% efficiency for normally distributed errors), both with no hard-cutoff step; time domain pre-whitening and tapering; and weighting the frequencies around each evaluation frequency using Parzen window weights. When using these additional steps individually, we find that the conclusions made regarding the differences each model's prediction ability do not change because the differences in each model's prediction performance are within the calculated uncertainty of the used method. Transfer function coefficients found using these modifications are generally within the error bars of those for the used procedure when the signal to noise ratio is above $\sim$3 (corresponding to periods higher than $\sim30$~s; At periods below $\sim$30~s, the electric field measurements are significantly influenced by the instrument response characteristics~\citep{Fujii2015}, aliasing, and the computed values are also expected to be significantly biased due to low signal to noise ratio.)

@@

The mean-squared-error on day $D$ is
\begin{equation*}
\mbox{MSE}_m = \frac{1}{N}\sum_{t=t_o}^{t_f} \big(GIC(t)-GIC_m(t)\big)^2 \,,
\end{equation*}
\noindent
where $GIC_m(t)$ is the model prediction of model $m$ given in Table~\ref{results} at time step $t$, $t_o=600$, and $t_f=85800$, so that sum is over one day of one-second-cadence measurements excluding the first and last 10~minutes.

The prediction efficiency, PE, is the ``Case I'' skill score described by \cite{Murphy1988} and is calculated for day $D$ using 
\begin{equation*}
\mbox{PE}_m = 1 - \frac{\sum_{t=t_o}^{t_f} \big(GIC(t)-GIC_m(t)\big)^2}{\sum_{t=t_o}^{t_f} \big(GIC(t)-\overline{GIC(t)}\big)^2}\,,
\end{equation*}
\noindent
where $\overline{GIC(t)}$ is an average over the same time interval. PE is a measure of the skill of the model relative to the variance in the data that are predicted and it represents the fraction of the variance in the data that is predicted by the model.

@@

The signal-to-noise ratio on day $D$ at a given evaluation frequency is the ratio of of the smoothed power spectrum value of $GIC$ to the smoothed power spectrum value of the prediction error at that evaluation frequency. The smoothed spectrum was obtained by averaging the raw spectrum in the same frequency windows used for the regression. The average signal-to-noise ratio at a given evaluation frequency is the average of the 35 out-of-sample signal-to-noise ratios at that frequency.

@@

Below $600$~s (10~min), Model~2 has the largest SN ratio. This result is somewhat visible in the sample interval shown in Figure~\ref{predictions} where the high-frequency fluctuations in the prediction error for Model~2 are the smallest among the models. For these periods, the predictive ability of each model is in the order that may be expected based on the data used to derive the model: Model~2, which uses direct measurements of the electric field, has a higher SN ratio than Model~3, which uses an indirect estimate of the electric field. Model~4 has a slightly higher SN than Model~3, and this is expected because they both use the same input of the magnetic field, but Model~3 has an additional source of uncertainty from the data-estimated coefficients in $\mathbf{Z}$.

@@

Above $10,000$~s (2.8~hrs), the SN ratio ordering is the same as the ordering of the PE and MSE ratios shown in Table~\ref{results}, with Model~1 having the lowest and Model~4 having the highest SN ratio.

@@

From Figure~\ref{Z} and Figure~\ref{Phi}, it follows that the 
In Figure~\ref{Phi}, at $T=12$~h, the phase difference between $a(\omega)$ and $b(\omega)$ of 180 degrees is the same as that for $a_o$ and $b_o$. In addition the ratio of $|b(\omega)|/|a(\omega)| =1.6$ is nearest to the frequency independent value predicted by Model~1 of $|\overline{b}_o|/|\overline{a}_o| = 1$.

Figure~\ref{Z_MT} and~\ref{Phi_MT} show the coefficients for $\mathbf{Z}$ in Model~3. Note that electric and magnetic field data from Memambetsu, Japan were also used by \cite{Fujii2015} to compute $\mathbf{Z}$. When using the \cite{Fujii2015} transfer function for Model~3, the PE was significantly lower (0.58 vs. 0.78). (The \cite{Fujii2015} transfer function was obtained from the IRIS DMC EMTF database [\cite{Kelbert2011}]). Although not critical to the issues addressed in this work, a preliminary investigation suggests that the cause is due to several factors. We have found that when using continuous one-month intervals prior to 2006 to compute $\mathbf{Z}$, our calculated transfer function matches the \cite{Fujii2015} transfer function to within the error bars except below periods of 100~s, which is likely explained by the issues with the electric field response in the 1-second-cadence measurements discussed in \cite{Fujii2015}. After 2006, our calculated transfer functions differ significantly from that in \cite{Fujii2015} - they are shifted downward by approximately 10-50\%. In addition, we have found that the estimated transfer function is sensitive to whether or not segment averaging was used. Due to the large gaps between and the varying length of continuous segments of data used in this work, transfer function coefficients were computed using one-day segments of data and then averaged. We have found that if a continuous set of measurements over approximately 21 days is used to compute a single transfer function, the transfer function magnitudes are up to 25\% higher than if 21 transfer functions were computed and then averaged. This result is unexpected, and extensive successful verification tests of our MT calculations along with the fact that this effect has observed by the co-author, who used different MT software on different data sets, suggests the explanation is not due to an error in the code used her to compute transfer functions. We expect to consider this in a future investigation as it suggests that stacking can be a significant source of transfer function bias and note that the most likely explanation involves non-stationarity in the noise.

@@

(3) The equation of Model~1 assumes that the electric field is spatially uniform at all points in the systems. The exact calculation of GICs in a power system under the quasi-DC assumption requires knowledge of the earth potential at the grounding points of the system, and simulations of GIC flow \citep{Albertson1981,Lehtinen1985} typically start with an assumed electric field along the lines to calculate these potential differences. The potential differences are then used along with the DC system component resistances and their connectivity to calculate GIC in the different parts of the system. In both Model~1 and Model~2, the input electric field is assumed to be spatially uniform along the line in which GIC was measured so that the potential difference is directly proportional to the electric field. Spatial variations in the electric field measured on Earth’s surface can be due to horizontal variations in the incident electric field and horizontal variations in sub-surface conductivity, including those at land-ocean interfaces. If a spatial variation exists along the line in which GIC was measured, the coefficients derived from the data in both models may differ significantly from those calculated using system configuration and resistance information. However, the data-derived coefficients will not have a frequency dependence that results from a variation in the direction of the electric field along the line.

@@

When GIC and magnetic field measurements are available, a model of the form of Model~4 can be used. For this particular data set, we have found that such a model provides the best estimates of GIC in terms of the PE and MSE metrics. Local ground magnetic field estimates along with GIC measurements could be used to derive a transfer function model that is updated when the system configuration changes and this model could be used to estimate GIC under extreme or simulated geomagnetic conditions. A key advantage of this approach is that long-term geomagnetic field measurements are generally easier to make and are more readily available. A disadvantage is the transfer function coefficients are difficult to interpret physically because they contain components of the transfer functions that connect the electric field to the magnetic field and GIC to the electric field and frequency-dependent artifacts due to violations of the assumption that the field is spectrally uniform.

@@

\begin{table}
\caption{Periods associated with evaluation frequency windows used for regression and smoothing spectra. \# is the evaluation frequency number, $N$ is the number of DFT points in the window, and $T_e=1/f_e$. The window range is $T_l$ through $T_h$.}
\centering
\begin{tabular}{l l l l l}
\hline \\
\# & $N$ & $T_e$ [s] & $T_l$ [s] & $T_h$ [s] \\
\hline \\
1 & 2 & 43200 & 28800 & 86400 \\
2 & 2 & 28800 & 21600 & 43200 \\
3 & 3 & 21600 & 14400 & 43200 \\
4 & 4 & 14400 & 9600 & 28800 \\
5 & 5 & 10800 & 7200 & 21600 \\
6 & 6 & 7854.5 & 5400 & 14400 \\
7 & 8 & 5760 & 3927.3 & 10800 \\
8 & 12 & 3927.3 & 2618.2 & 7854.5 \\
9 & 16 & 2880 & 1920 & 5760 \\
10 & 22 & 2009.3 & 1350 & 3927.3 \\
11 & 31 & 1440 & 960 & 2880 \\
12 & 43 & 1016.5 & 680.3 & 2009.3 \\
13 & 61 & 720 & 480 & 1440 \\
14 & 85 & 511.2 & 341.5 & 1016.5 \\
15 & 120 & 361.5 & 241.3 & 720 \\
16 & 170 & 255.6 & 170.4 & 511.2 \\
17 & 240 & 180.8 & 120.5 & 361.5 \\
18 & 338 & 128 & 85.4 & 255.6 \\
19 & 478 & 90.5 & 60.3 & 180.8 \\
20 & 676 & 64 & 42.7 & 128 \\
21 & 956 & 45.2 & 30.2 & 90.5 \\
22 & 1351 & 32 & 21.3 & 64 \\
23 & 1910 & 22.6 & 15.1 & 45.2 \\
24 & 2701 & 16 & 10.7 & 32 \\
25 & 3819 & 11.3 & 7.5 & 22.6 \\
26 & 5401 & 8 & 5.3 & 16 \\
27 & 7638 & 5.7 & 3.8 & 11.3 \\
28 & 10801 & 4 & 2.7 & 8 \\
\hline \\
\end{tabular}
\label{evaluationperiods}
\end{table}

>>>>

Motivated by the results described in the following section, in which Model~2 provides a significantly better representation of GIC than Model~1, we have also considered two additional models that can be used to specify GIC. In these two models, a direct measurement of the geoelectric field is not used; instead, the driver is the geomagnetic field, which is much more frequently measured, and as a result, these models may be of practical interest.

@@

Estimates of $a(\omega)$ and $b(\omega)$ are made at evaluation frequencies, $\omega_e \equiv 2\pi f_e$, by performing a ordinary least squares regression (OLS) using Equation~\ref{model2} and a set $G_E(\omega)$ and $\mathbf{E}(\omega)$ in a band of frequencies around $f_e$ from the DFT (discrete Fourier transform) of 24 hours of 1-s cadence data. Following the suggestion of \cite{Simpson2005}, the 28 values of $f_e$ start at $0.25$ Hz and decrease by a factor of approximately $\sqrt{2}$ and end at $1/43200$~Hz; if a $f_e$ value computed in this way is not an integer multiple of $1/86400$, the nearest smaller integer multiple is used. The periods associated with the evaluation frequencies are listed in Table~\ref{evaluationperiods}. The regression for each $f_e$ uses all DFT frequencies with values in a window with a width of twice the separation from the next lower value of $f_e$ and the given $f_e$ with the exceptions that the lowest evaluation frequency band has a lower bound of $1/86400$~Hz and highest evaluation frequency band has an upper bound of $0.375$~Hz.

@@

Several modifications to this basic procedure were considered. Robust regression \citep{Egbert1986} with a Huber weighting constant of $1.345$ and an upper limit of 50 iterations and bi-square weighting with a constant of $4.685$ (both constants provide 95\% efficiency for normally distributed errors), both with no hard-cutoff step; time domain pre-whitening and tapering; and weighting the frequencies around each evaluation frequency using Parzen window weights. When using these additional steps individually, we find that the conclusions made regarding the differences each model's prediction ability do not change because the differences in each model's prediction performance are within the calculated uncertainty of the used method. Transfer function coefficients found using these modifications are generally within the error bars of those for the used procedure when the signal to noise ratio is above $\sim$3 (corresponding to periods longer than $\sim30$~s; At periods below $\sim$30~s, the electric field measurements are significantly influenced by the instrument response characteristics~\citep{Fujii2015}, aliasing, and the computed values are also expected to be significantly biased due to low signal to noise ratio.)

@@

The mean-squared-error for model $m$ on day $D$ is
\begin{equation*}
\mbox{MSE}_m = \frac{1}{N}\sum_{t=t_o}^{t_f} \big(GIC(t)-GIC_m(t)\big)^2 \,,
\end{equation*}
\noindent
where $GIC_m(t)$ is the model prediction of model $m$ given in Table~\ref{results} at time step $t$, $t_o=600$, and $t_f=85800$, so that sum is over one day of one-second-cadence measurements excluding the first and last 10~minutes.

The prediction efficiency, PE, is the ``Case I'' skill score described by \cite{Murphy1988} and is calculated for day $D$ using 
\begin{equation*}
\mbox{PE}_m = 1 - \frac{\sum_{t=t_o}^{t_f} \big(GIC(t)-GIC_m(t)\big)^2}{\sum_{t=t_o}^{t_f} \big(GIC(t)-\overline{GIC(t)}\big)^2}\,,
\end{equation*}
\noindent
where $\overline{GIC(t)}$ is an average over the same time interval used in the summations. The PE is a measure of the skill of the model relative to the variance in the data that are predicted and it represents the fraction of the variance in the data that is predicted by the model.

@@

The signal-to-noise ratio on day $D$ at a given evaluation frequency is the ratio of of the smoothed power spectrum value of $GIC$ to the smoothed power spectrum value of the prediction error at that evaluation frequency. The smoothed spectrum was obtained by averaging the raw spectrum in the same frequency bands used for the regression. The average signal-to-noise ratio at a given evaluation frequency is the average of the 35 out-of-sample signal-to-noise ratios at that frequency.

@@

Below $600$~s (10~min), Model~2 has the largest SN ratio. This result is somewhat visible in the sample interval shown in Figure~\ref{predictions} where the high-frequency fluctuations in the prediction error for Model~2 are the smallest among the models. For these periods, the predictive ability of each model is in the order that may be expected based on the data used to derive the model: Model~2, which uses direct measurements of the electric field, has a larger SN ratio than Model~3, which uses an indirect estimate of the electric field. Model~4 has a slightly larger SN than Model~3, and this is expected because they both use the same input of the magnetic field, but Model~3 has an additional source of uncertainty from the data-estimated coefficients in $\mathbf{Z}$.

@@

Above $10,000$~s (2.8~hrs), the SN ratio ordering is the same as the ordering of the PE and MSE ratios shown in Table~\ref{results}, with Model~1 having the lowest and Model~4 having the largest SN ratio.

@@

%From Figure~\ref{Z} and Figure~\ref{Phi}, it follows that the In Figure~\ref{Phi}, at $T=12$~h, the phase difference between $a(\omega)$ and $b(\omega)$ of 180 degrees is the same as that for $a_o$ and $b_o$. In addition the ratio of $|b(\omega)|/|a(\omega)| =1.6$ is nearest to the frequency independent value predicted by Model~1 of $|\overline{b}_o|/|\overline{a}_o| = 1$.

Figure~\ref{Z_MT} and~\ref{Phi_MT} show the parameters for $\mathbf{Z}$ in Model~3. Note that electric and magnetic field data from Memambetsu, Japan were also used by \cite{Fujii2015} to compute $\mathbf{Z}$. When using the \cite{Fujii2015} transfer function for Model~3, the PE was significantly lower (0.58 vs. 0.78). (The \cite{Fujii2015} transfer function was obtained from the IRIS DMC EMTF database [\cite{Kelbert2011}]). Although not critical to the issues addressed in this work, a preliminary investigation suggests that the cause is due to several factors. We have found that when using continuous one-month intervals prior to 2006 to compute $\mathbf{Z}$, our calculated transfer function matches the \cite{Fujii2015} transfer function to within the error bars except below periods of 100~s, which is likely explained by the issues with the electric field response in the 1-second-cadence measurements discussed in \cite{Fujii2015}. After 2006, our calculated transfer functions differ significantly from that in \cite{Fujii2015} - they are shifted downward by approximately 10-50\%. In addition, we have found that the estimated transfer function is sensitive to whether or not segment averaging was used. Due to the large gaps between and the varying length of continuous segments of data used in this work, transfer function coefficients were computed using one-day segments of data and then averaged. We have found that if a continuous set of measurements over approximately 21 days is used to compute a single transfer function, the transfer function magnitudes are up to 25\% larger than if 21 transfer functions were computed and then averaged. This result is unexpected, and extensive successful verification tests of our MT calculations along with the fact that this effect has observed by the co-author, who used different MT software on different data sets, suggests the explanation is not due to an error in the code used her to compute transfer functions. We expect to consider this in a future investigation as it suggests that stacking can be a significant source of transfer function bias and note that the most likely explanation involves non-stationarity in the noise.

@@

(3) The equation of Model~1 assumes that the electric field is spatially uniform at all points in the systems. (An equation of this form can be computed in principle for an electric field with an arbitrary spatial distribution but most often the electric field is assumed to be spatially uniform). The exact calculation of GICs in a power system under the quasi-DC assumption requires knowledge of the electric field at all points of the system, and simulations of GIC flow typically start with an assumed electric field along the lines to calculate input voltages \citep{Albertson1981,Lehtinen1985, Horton2012}. The voltages are then used along with the DC system component resistances and their connectivity to calculate GIC in the different parts of the system. In both Model~1 and Model~2, the input electric field is assumed to be spatially uniform along the line in which GIC was measured so that the potential difference is directly proportional to the electric field. Spatial variations in the electric field measured on Earth's surface can be due to horizontal variations in the incident electric field and horizontal variations in sub-surface conductivity, including those at land-ocean interfaces. If a spatial variation exists along the line in which GIC was measured, the coefficients derived from the data in both models may differ significantly from those calculated using system configuration and resistance information. However, the data-derived coefficients will not have a frequency dependence that results from a variation in the direction of the electric field along the line.

@@

When GIC and magnetic field measurements are available, a model of the form of Model~4 can be used. For this particular data set, we have found that such a model provides the best estimates of GIC in terms of the PE and MSE metrics. Local ground magnetic field estimates along with GIC measurements could be used to derive a transfer function model that is updated when the system configuration changes and this model could be used to estimate GIC under extreme or simulated geomagnetic conditions. A key advantage of this approach is that long-term geomagnetic field measurements are generally easier to make and are more readily available. A disadvantage is the transfer function parameters are difficult to interpret physically because they contain components of the transfer functions that connect the electric field to the magnetic field and GIC to the electric field and frequency-dependent artifacts due to violations of the assumption that the field is spectrally uniform.

@@

\begin{table}
\caption{Periods associated with evaluation frequency bands used for regression and smoothing spectra. \# is the evaluation frequency number, $N$ is the number of DFT points in the window, and $T_e=1/f_e$. The window range is $T_l$ through $T_h$.}
\centering
\begin{tabular}{l l l l l}
\hline \\
\# & $N$ & $T_e$ [s] & $T_l$ [s] & $T_h$ [s] \\
\hline \\
1 & 2 & 43200 & 28800 & 86400 \\
2 & 2 & 28800 & 21600 & 43200 \\
3 & 3 & 21600 & 14400 & 43200 \\
4 & 4 & 14400 & 9600 & 28800 \\
5 & 5 & 10800 & 7200 & 21600 \\
6 & 6 & 7854.5 & 5400 & 14400 \\
7 & 8 & 5760 & 3927.3 & 10800 \\
8 & 12 & 3927.3 & 2618.2 & 7854.5 \\
9 & 16 & 2880 & 1920 & 5760 \\
10 & 22 & 2009.3 & 1350 & 3927.3 \\
11 & 31 & 1440 & 960 & 2880 \\
12 & 43 & 1016.5 & 680.3 & 2009.3 \\
13 & 61 & 720 & 480 & 1440 \\
14 & 85 & 511.2 & 341.5 & 1016.5 \\
15 & 120 & 361.5 & 241.3 & 720 \\
16 & 170 & 255.6 & 170.4 & 511.2 \\
17 & 240 & 180.8 & 120.5 & 361.5 \\
18 & 338 & 128 & 85.4 & 255.6 \\
19 & 478 & 90.5 & 60.3 & 180.8 \\
20 & 676 & 64 & 42.7 & 128 \\
21 & 956 & 45.2 & 30.2 & 90.5 \\
22 & 1351 & 32 & 21.3 & 64 \\
23 & 1910 & 22.6 & 15.1 & 45.2 \\
24 & 2701 & 16 & 10.7 & 32 \\
25 & 3819 & 11.3 & 7.5 & 22.6 \\
26 & 5401 & 8 & 5.3 & 16 \\
27 & 7638 & 5.7 & 3.8 & 11.3 \\
28 & 10801 & 4 & 2.7 & 8 \\
\hline \\
\end{tabular}
\label{evaluationperiods}
\end{table}
